---
title: "ESM 244 | Lab 7 | Spatial Extravaganza"
author: "Hanna Buechi"
date: "2/20/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

library(tidyverse)
library(tmap)
library(sf)
library(spatstat)
library(maptools)
library(sp)
library(raster)
# library(rgdal)
library(gstat)

```

Part 1. Hawaii raster practice
```{r}

# read in radiation raster data, Hawaii
hi_par <- raster("PAR_CLIM_M.tif")
plot(hi_par)

hi_sst <- raster("SST_LTM.tif")
plot(hi_sst)

hi_chl <- raster("CHL_LTM.tif")
plot(hi_chl)

par(mfrow = c(1,3))
plot(hi_par)
plot(hi_sst)
plot(hi_chl)

# in console
# hi_sst
# class       : RasterLayer 
# dimensions  : 801, 1217, 974817  (nrow, ncol, ncell)
# resolution  : 500, 500  (x, y)
# extent      : 351124.5, 959624.5, 2079231, 2479731  (xmin, xmax, ymin, ymax)
# coord. ref. : +proj=utm +zone=4 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 
# data source : /Users/hannabuechi/github/esm-244-lab-7/SST_LTM.tif 
# names       : SST_LTM

# more info: hi_sst@crs or hi_sst@extent
```

Reproject to WGS84:
```{r}

# spacing matters
# flattening function ellps

# save and copy this to use it a lot
wgs84 <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +no_defs"

hi_sst_84 <- projectRaster(hi_sst, crs = wgs84, method = "bilinear")

hi_sst_84@crs # output:

#CRS arguments:
 # +proj=longlat +datum=WGS84 +ellps=WGS84 +no_defs +towgs84=0,0,0

hi_sst_84@extent # values line up with lat/long values

plot(hi_sst_84)

```

raster::aggregate() for resampling
```{r}

# aggregate cells via resampling so that resolution is lower and it's easier to run and visualize; get code running instead of needing to wait

sst_rs <- aggregate(hi_sst, fact = 10)

par(mfrow=c(1,2))
plot(hi_sst)
plot(sst_rs)



```

Crop a raster:
```{r}

hi_sst_84@extent
plot(hi_sst_84)

bounds <- as(extent(-156.2, -154.5, 18.7, 20.4), 'SpatialPolygons') # as() is a general wrapper that says take this thing and turn it into this thing # Allison tested the lats and longs of extent() values, it wasn't magic --> see x and y axes of hi_sst_84 raster
# doesn't have a coordinate reference system yet

crs(bounds) <- crs(hi_sst_84) # crs() only works if a raster doesn't have a crs associated with it, otherwise have to use projectRaster() to change it

sst_crop <- crop(hi_sst_84, bounds) # crop the raster by these bounds
plot(sst_crop)

```

Simple raster math:


Non-sensical variable called "tropicality" that is the sum of the PAR + SST + 2*ChlA, and we want to map that variable...


Note: for thematic (categorical) raster data, consider using method = "ngm"
```{r}

# need to reclassify the other crs too
hi_par_84 <- projectRaster(hi_par, crs = wgs84, method = "bilinear")

hi_chla_84 <- projectRaster(hi_chl, crs = wgs84, method = "bilinear")

par(mfrow = c(1,3))
plot(hi_sst_84)
plot(hi_par_84)
plot(hi_chla_84)

# scales are different
```

```{r}

trop <- hi_par_84 + hi_sst_84 + 2*hi_chla_84

# message: Raster objects have different extents. Result for their intersection is returned --> if there are cells that show up in one raster and not the others, those won't be plotted

plot(trop)

```

Now let's try to look at something using tmap():

```{r}

islands <- read_sf(dsn = 'islands', layer = "Island_boundaries") %>% 
  dplyr::select(Island) %>%  # because sf, it will keep geometries
  st_simplify(dTolerance = 10) %>% # did we use this in Lab 6?
  st_transform(crs = 4326) # re-project

plot(islands)

```

```{r}

# tmap can be interactive with tmap_mode("view")

tmap_mode("plot") # but we want static

sst_map <- tm_shape(hi_sst_84) + # not the tropicality one =)
  tm_raster(title = "Mean Sea Surface Temperature") +
  tm_layout(bg.color = "navyblue",
            legend.position = c("left", "bottom"),
            legend.text.color = "white",
            legend.text.size = 0.5) +
  tm_shape(islands) + # add the shape file
  tm_fill("darkgreen") # fill in the land

# notice that the left most islands doesn't exist in our shapefile plot(islands)

tmap_save(sst_map, "hanna_hawaii.png", height = 5) # this shows up in my working directory

```

Conditional rasters and masking

Let's say: we have a sensitive species and we're trying to find habitat that it might like. We know: they like warm water (average SST >= 25.6 degrees) and solar radiation (PAR) below 54. (Quite simplified.)

```{r}

# we will need these rasters to line up

hi_sst_84

#class       : RasterLayer 
#dimensions  : 822, 1241, 1020102  (nrow, ncol, ncell)
#resolution  : 0.0048, 0.00452  (x, y)
#extent      : -160.4705, -154.5137, 18.7309, 22.44634  (xmin, xmax, ymin, ymax)
#coord. ref. : +proj=longlat +datum=WGS84 +ellps=WGS84 +no_defs +towgs84=0,0,0 
#data source : in memory
#names       : SST_LTM 
#values      : 25.05634, 26.05227  (min, max)

hi_par_84

#class       : RasterLayer 
#dimensions  : 822, 1229, 1010238  (nrow, ncol, ncell)
#resolution  : 0.0048, 0.00452  (x, y)
#extent      : -160.4365, -154.5373, 18.7309, 22.44634  (xmin, xmax, ymin, ymax)
#coord. ref. : +proj=longlat +datum=WGS84 +ellps=WGS84 +no_defs +towgs84=0,0,0 
#data source : in memory
#names       : PAR_CLIM_M 
#values      : 36.2316, 56.4165  (min, max)

# dimensions (columns and rows) and extent are different

```

```{r}

extent(hi_sst_84) <- extent(hi_par_84)
cr <- raster(nrow = 822, # we want this raster to match hi_par_84 via copy and paste --> making this isn't necessary
             ncol = 1229,
             xmn = -160.4365,
             xmx = -154.5373,
             ymn = 18.7309,
             ymx = 22.44364) 

# resample hi_sst_84 to match hi_par_84

sst_new <- resample(hi_sst_84, hi_par_84, method = "bilinear")

# OR

sst_new_cr <- resample(hi_sst_84, cr, method = "bilinear") # math is slightly different than sst_new --> think about why with the bilinear interpolation

compareRaster(sst_new, hi_par_84) # TRUE: they are matching

plot(sst_new) # everything is filled in now
plot(hi_par_84)

# before resampling:
# compareRaster(hi_sst_84, hi_par_84)
# Error in compareRaster(hi_sst_84, hi_par_84) : different number or columns

```

Make a cropped version just for Kauai:
```{r}

bounds_main <- as(extent(-159.9, -159.2, 21.7, 22.3), 'SpatialPolygons')
crs(bounds_main) <- crs(sst_new)

par_kauai <- crop(hi_par_84, bounds_main)
sst_kauai <- crop(sst_new, bounds_main)

plot(par_kauai)
plot(sst_kauai) # the empty cells are filled in


```


Now we only want to isolate regions where temp >= 25.4, PAR < 54
```{r}

par_hab <- par_kauai
par_hab[par_hab >= 54] <- NA # base R notation because sf() can't use tidy format in this case
plot(par_hab) # only areas according to our restrictions show up

sst_hab <- sst_kauai
sst_hab[sst_hab < 25.4] <- NA
plot(sst_hab)

# find where these rasters overlap via masking, raster::mask

suit_hab <- mask(sst_hab, par_hab) # I imagine that we could have done the other way around
plot(suit_hab)

```

####Part 2. Point pattern analysis

An analysis of red tree voles in Humboldt County

```{r}

voles <- read_sf(dsn = 'redtreevoledata', layer = "ds033") %>% 
  dplyr::select(COUNTY) %>% 
  filter(COUNTY == "HUM") %>% 
  st_transform(crs = 4326)

plot(voles)

# Get the Humboldt County outline

humboldt <- read_sf(dsn = 'redtreevoledata', layer = "california_county_shape_file") %>% 
  filter(NAME == "Humboldt") %>% 
  dplyr::select(NAME)

# no .prj file in california_county_shape_file

st_crs(humboldt) <- 4326

plot(humboldt)

tm_shape(humboldt) +
  tm_fill() +
  tm_shape(voles) +
  tm_dots(size = 0.2)

ggplot() +
  geom_sf(data = humboldt) +
  geom_sf(data = voles)

ggsave("humvoles.png",
       units = "in",
       width = 4,
       height = 6,
       dpi = 300)

# is there a clustering pattern?



```


Sf hasn't caught up with point pattern analysis, so we will have to convert some objects from sf-based objects to general spatial objects and then point pattern objects.

We want to explore point patterns a few different ways:

- quadrat analysis
- distance based (neighbor analysis, G-function and K-function)

```{r}

voles_sp <- as(voles, 'Spatial') # general spatial
voles_ppp <- as(voles_sp, "ppp") # ppp is a point pattern analysis object

# county boundary will be the spatial window

humboldt_sp <- as(humboldt, "Spatial")
humboldt_win <- as(humboldt_sp, "owin") # outer window

voles_pb <- ppp(voles_ppp$x, voles_ppp$y, window = humboldt_win) # lat and long and window

```

Quadrat test:
```{r}

voles_qt <- quadrat.test(voles_pb, nx = 5, ny = 10)

# Testing the null hypothesis of spatial evenness (although you'll hear it called a test for CSR)

voles_qt

#Chi-squared test of CSR using quadrat counts
#	Pearson X2 statistic

#data:  voles_pb
#X2 = 388.72, df = 44, p-value < 2.2e-16
#alternative hypothesis: two.sided

#Quadrats: 45 tiles (irregular windows)

# there's a very low p-value so REJECT NULL
# we would conclude that these events do NOT reflect spatial evenness

plot(voles_pb)
plot(voles_qt, add = TRUE, cex = 0.4)

# top left: number of observations, top right: events expected if true CSR, other number: metric of above/below intensity

```

Plotting kernel densities for spatial data:

```{r}

point_density <- density(voles_pb, sigma = 0.02) # sigma = bandwidth (remember from lecture notes) --> a lot of power
plot(point_density)

# kernel densities can be great for showing hot spots but be very, very careful

vole_raster <- raster(point_density, crs = wgs84)

tm_shape(vole_raster) +
  tm_raster()

```

Nearest neighbor

G-function: considers the distance of each observation to its NEAREST neighbor

K-function: considers how close all neighboring observations are to an event (concentric circles)

```{r}

r <- seq(0, 0.15, by = 0.005) # making the seqence of distances overwhich calculate nearest-neighbor distances # units of the CRS

gfunction <- envelope(voles_pb, fun = Gest, r = r, nsim = 20) # simulations of the G-function if there was true CSR + nsim = number of MC simulations

plot(gfunction$obs ~ gfunction$r, type = "l", col = "black") # clear why Allison chose those increments for r
lines(gfunction$theo ~ gfunction$r, type = "l", col = "red") # run these two lines at the same time

# nearest neighbors are closer than what we expected

#K/L function: Asking about how close ALL neighbors are to EVERY event in the spatial window

r2 <- seq(0, 0.5, by= 0.05) # need much larger space to consider multiple neighbors

lfunction <- envelope(voles_pb, fun = Lest, r = r2, nsim = 20, global = TRUE) # this takes a lot longer than G-function

plot(lfunction$obs ~ lfunction$r, type = "l", col = "blue")
lines(lfunction$theo ~ lfunction$r, type = "l", col = "red")

# lower distances, greater density of events than for CSR, indicative of clustering

```


Diggle-Cressie-Loosemore-Ford test of CSR
```{r}

DCLFTest <- dclf.test(voles_pb, nsim = 30)
DCLFTest

# p-value = 0.03226 This is different than expected from CSR
# this is not immune to the same problems of other statistical test - if there is increased power, you will see something; if there are few events, you won't see anything ever if there is significant difference
```

###Part 3. Spatial interpolation by krieging


```{r}

ks_rain <- read_csv("KSRain2.csv") # not inherintly spatial

ks_sf <- st_as_sf(ks_rain, coords = c("LON", "LAT"), crs = 4326) # must be in long lat order
plot(ks_sf)

ks_counties <- read_sf(dsn = 'KSCounties', layer = "ks_counties_shapefile")
st_crs(ks_counties)

plot(ks_counties)

tm_shape(ks_counties) +
  tm_fill() +
  tm_shape(ks_sf) +
  tm_dots("AMT", size = 0.5)

# there are some pretty large areas of the state where there is no data, so a little worry for predictions (there will be higher error up there)


```

```{r}


 # kriging also wants a different kind of object

ks_sp <- as_Spatial(ks_sf)


```


Make a spatial grid to interpolate values over
```{r}

lat <- seq(37, 40, length.out = 200)
long <- seq(-94.6, -102, length.out = 200)

grid <- expand.grid(lon = long, lat = lat) # creates a spatial grid
grid_sf <- st_as_sf(grid, coords = c("lon", "lat"), crs = 4326)
grid_sp <- as_Spatial(grid_sf)

```

Then make a variogram and find the variogram model
```{r}

ks_vgm <- variogram(AMT ~ 1, ks_sp) # ~ 1, three types of kriging: this is ordinary kriging
plot(ks_vgm)

ks_vgm_fit <- fit.variogram(ks_vgm, model = vgm(nugget = 0.2, psill = 0.8, range = 200, model = "Sph"))

plot(ks_vgm, ks_vgm_fit)

```

Now we can do spatial interpolation:
```{r}

ks_krige <- krige(AMT ~ 1, ks_sp, grid_sp, model = ks_vgm_fit)

# make this a dataframe

ks_krige_df <- as.data.frame(ks_krige)

# long, lat, rainfall prediction, variance

ks_krige_2 <- ks_krige_df %>% 
  rename(lon = coords.x1, lat = coords.x2, predicted = var1.pred, err = var1.var)

rain_predicted <- st_as_sf(ks_krige_2, coords = c("lon", "lat"), crs = 4326)

# outline of kansas and crop to that

ks <- read_sf(dsn = "states", layer = "cb_2017_us_state_20m") %>% 
  dplyr::select(NAME) %>% 
  filter(NAME == "Kansas") %>% 
  st_transform(crs = 4326)

```





























